---
title: "TC data analysis -- SI and SP basins"
author: "Melissa Renard"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, warning=FALSE, include=FALSE}
rm(list = ls())

# ChatGPT used for the below code: required packages
# Define a list of required packages
required_packages <- c("tidyverse", "grid", "gridExtra", "corrplot")

# Check if the packages are installed, and if not, install them
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}
# Load the packages into the session
invisible(lapply(required_packages, library, character.only = TRUE))

knitr::opts_chunk$set(echo = TRUE) 
```

```{r basin-choice}
# CHOOSE BASIN HERE -- 'SI' OR 'SP'.
basin_choice <- 'SI'
  
if (basin_choice == 'SI'){
  x_limits <- c(30, 140)
  x_breaks <- seq(30,140,10)
  y_limits <- c(-40, 0)
  y_breaks <- seq(-40,0,10)
} else if (basin_choice == 'SP'){
  x_limits <- c(130, 240)
  x_breaks <- seq(130,240,10)
  y_limits <- c(-40, 0)
  y_breaks <- seq(-40,0,10)
} else {
  stop("basin_choice must be 'SI' or 'SP'.")
}
```

```{r}
# Load tracks data
tracks <- read.csv("../Data/tracks.csv", na="") %>% as_tibble %>% 
  filter(basin == basin_choice)

tracks$category <- tracks$category %>% as.factor
tracks <- tracks %>% rename('t_day' = 'days_since_1980', 
                  't_year' = 'years_since_1980', 
                  'Nt' = 'TC_num') %>%
  mutate(t_month = lapply(t_year, function(x) x*12) %>% unlist,
         east170 = lon > 170,
         east75 = lon > 75,
         range_sw = max_sw - min_sw,
         Nt = as.integer(rownames(tracks)))

# TCs in the southern hemisphere occur between November and April. 
# Need to write a new dataframe for that.
tracks_1 <- tracks %>% mutate(enso_year = year + (month > 7)) %>%
  filter(enso_year >= 1981 & enso_year <= 2023)
```

```{r}
# Load ENSO data
enso_phases <- c("Nina", "Neutral", "Nino")
enso_labels <- c("La Niña", "Neutral", "El Niño")

enso_df <- read.csv('../Data/ENSO.csv', skip=3) %>% as_tibble %>%
  mutate(enso = factor(enso, levels = enso_phases))
enso_df_TC <- enso_df %>% filter(year >= 1981 & year <= 2023)

tracks_1 <- tracks_1 %>% left_join(enso_df, by = join_by(enso_year==year))

```

```{r}
ggplot(tracks_1, aes(x = lon, y = lat, color = enso)) +
  geom_point(data = subset(tracks_1, enso== 'Neutral'), aes(color = enso), size = 1.5)+
  geom_point(data = subset(tracks_1, enso!= 'Neutral'), aes(color = enso), size = 1.5) +
  scale_x_continuous(limits = x_limits, breaks = x_breaks, expand = c(0,1)) +
  scale_y_continuous(limits = y_limits, breaks = y_breaks, expand=c(0,1)) +
  labs(
    title = "Tropical cyclone genesis in the SI basin 1980-2023",
    x = "Longitude (\u00B0E)",
    y = "Latitude (\u00B0)",
    color = "ENSO Phase"
  ) +
  coord_equal() +  # Ensure equal scaling of x and y axes
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    legend.position = 'bottom',
  ) +
  scale_color_manual(
    name = "",
    breaks = enso_phases,
    values = c("red", "grey", "blue"),
    labels = enso_labels
  )
```

```{r}
p1 <- ggplot(tracks, aes(x = year)) +
  geom_bar() + 
  theme_classic() + 
  labs(x = "Year", y ="Count", title = "Number of TCs in SP basin\nby year") + 
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(tracks_1, aes(x = month)) + 
  geom_bar() +
  theme_classic() +
  labs(x = "Month", y = "Count", title = "Number of TCs in SP basin\nby month") + 
  theme(plot.title = element_text(hjust = 0.5))

p3 <- ggplot(tracks_1, aes(x = category)) + 
  geom_bar() +
  theme_classic() +
  labs(x = "Month", y = "Count", title = "Number of TCs in SP basin\nby Saffir-Simpson category") + 
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(p1,p2,p3, ncol=2, nrow=2)
```

# Frequency
## Frequency based on ENSO phases

```{r}
my_lm <- glm(table(tracks_1$enso_year) ~ 1, family = poisson)
summary(my_lm)
mu <- my_lm$coefficients[[1]]
exp(mu)
```

```{r} 
my_lm <- glm(table(tracks_1$enso_year) ~ enso_df_TC$enso, family = poisson)

summary(my_lm)

mu <- c(my_lm$coefficients[[1]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[2]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[3]])

exp(mu)
```
Estimated mean frequency of TCs in SI basin in La Niña, neutral and El Niño phases, respectively: `r round(exp(mu),2)`

```{r, fig.width = 3.5, fig.height=4}
tracks_1 %>% 
  count(enso) %>% 
  mutate(fqcy = n/c(9,24,10),
         sde = summary(my_lm)$coef[,2] %>% as.vector) %>% 
  ggplot(aes(x = enso, y = fqcy)) + 
  geom_bar(stat = 'identity', fill = 'grey') + 
  geom_errorbar(aes(x =enso, ymin = fqcy*exp(1.96*sde), 
                    ymax = fqcy*exp(-1.96*sde)), width = 0.3, color="blue", 
                size=0.8)+
  theme_minimal() + 
  labs(x = 'ENSO phase', y = 'Average annual frequency') + 
  scale_x_discrete(
    labels = enso_labels
  ) +
  theme(legend.position = 'None')
```

# Locations frequency based on ENSO phases

```{r}
summary(tracks_1$lon)
summary(tracks_1$lat)

my_lm <- lm(lon ~ enso, data=tracks_1)
summary(my_lm)

mu <- c(my_lm$coefficients[[1]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[2]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[3]])

mu
```
Estimated mean longitude of TCs in La Niña, neutral and El Niño phases, respectively: `r round(mu,3)`.

This is significant. La Niña means more TCs close to Australia.

```{r}
if (basin_choice == 'SI'){
  my_lm <- glm(east75 ~ enso, data=tracks_1, family = binomial)
} else {
  my_lm <- glm(east170 ~ enso, data=tracks_1, family = binomial)
}

summary(my_lm)

mu <- c(my_lm$coefficients[[1]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[2]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[3]])
exp(mu)/(1+exp(mu))
```
Estimated probability of a TC produced east of $75^\circ$E in La Niña, neutral and El Niño phases, respectively: `r exp(mu)/(1+exp(mu))`

# Intensity
## Intensity based on ENSO phases

```{r, fig.height = 4, fig.width=3.5}
my_lm <- glm(max_sw ~ enso, data=tracks_1, family = Gamma(link=log))

summary(my_lm)

mu <- c(my_lm$coefficients[[1]],
        my_lm$coefficients[[1]] + my_lm$coefficients[[2]],
        my_lm$coefficients[[1]] + my_lm$coefficients[[3]])

exp(mu)
sde <- summary(my_lm)$coef[,2] %>% as.vector

tracks_1 %>% 
  count(enso) %>% 
  mutate(sev = exp(mu),
         sde = summary(my_lm)$coef[,2] %>% as.vector) %>% 
  ggplot(aes(x = enso, y = sev)) + 
  geom_bar(stat = 'identity', fill = 'grey') + 
  geom_errorbar(aes(x =enso, ymin = exp(mu - 1.96*sde), 
                    ymax = exp(mu + 1.96*sde)), width = 0.3, color="blue", 
                size=0.8)+
  theme_minimal() + 
  labs(x = 'ENSO phase', y = 'Intensity (knots)') + 
  scale_x_discrete(
    labels = enso_labels
  ) +
  theme(legend.position = 'None')
```
The estimated mean maximum wind speed of a TC in La Niña, Neutral and El Niño phases are, respectively: `r round(1/mu,2)`.

```{r}
p1 <- ggplot(data = tracks_1, aes(x = category, y = range_sw)) + 
  geom_violin(aes(fill = category)) + 
  labs(title = "Wind speed range by category") + 
  theme(legend.position = "none")

p2 <- ggplot(data = tracks_1, aes(x = max_sw, y = range_sw)) + 
  geom_point() + 
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

# Track lifetime

```{r}
my_lm <- glm(lifetime ~ enso, data = tracks_1, family=Gamma(link=log))
summary(my_lm)

mu <- c(my_lm$coefficients[[1]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[2]], 
        my_lm$coefficients[[1]] + my_lm$coefficients[[3]])

exp(mu)*365

p1 <- ggplot(data = tracks_1, aes(x = lifetime, y = after_stat(density))) + 
  geom_freqpoly(aes(color= enso), bins=20) + 
  theme_minimal() + 
  theme(legend.position = 'bottom') +
  labs(title = "Track lifetime by ENSO phase") + 
  scale_color_manual(
    name = "ENSO phase",
    labels = enso_labels,
    values = c("red", "grey50", "blue")
  )

p2 <- ggplot(data = tracks_1, aes(x = enso, y = lifetime)) + 
  geom_violin(aes(fill = enso)) + 
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(title = "Track lifetime by ENSO phase") + 
  scale_fill_manual(
    name = "ENSO phase",
    labels = enso_labels,
    values = c("red", "grey50", "blue")
  ) +
  scale_x_discrete(labels = enso_labels)

grid.arrange(p1,p2, ncol = 2)
```

```{r}
(cor_table <- cor(tracks_1 %>% select(max_sw, lat, lon, lifetime)))
corrplot(cor_table, method = "circle")
```
